---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the data from "activity.csv".

```{r, activity_data}
activity_data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
Calculates the total number of steps taken per day and creates a histogram steps taken each day. The calculates the mean and median of the total number of steps taken per day. 

```{r, daily_steps, warning=FALSE, message=FALSE}
library(dplyr)
activity_summary <- activity_data %>% select(steps, date) %>% group_by(date) %>% summarise(total_steps = sum(steps, na.rm = TRUE))
hist(activity_summary$total_steps, xlab = "Total Daily Steps", main = "Histogram of Daily Step Totals")
daily_mean <-mean(activity_summary$total_steps, na.rm = TRUE)
daily_median <-median(activity_summary$total_steps, na.rm = TRUE)

```

The average number of steps taken per day is `r format(daily_mean, scientific = FALSE)` and the median is `r format(daily_median, scientific = FALSE)`.


## What is the average daily activity pattern?
```{r, activity_patter}
interval_summary <- activity_data %>% select(steps, date, interval) %>% group_by(interval) %>% summarise(average_steps = mean(steps, na.rm=TRUE))
plot(interval_summary$interval, interval_summary$average_steps, type = 'l', ylab = "Average Daily steps", xlab = "Interval")
most_steps <- interval_summary[interval_summary$average_steps == max(interval_summary$total_steps),1]
most_steps <- most_steps$interval[1]
```

The time interval with the maximum number is: ```r most_steps```.

## Inputing Missing Values
Missing vaues will be replaces with the average value for that interval rounded to the nearest step. 
```{r, missing_values, warning=FALSE, message=FALSE}
total_missing <- nrow(activity_data[is.na(activity_data$steps),])


replace_data <- activity_data
interval_averages <- replace_data %>% select(steps, date, interval) %>% group_by(interval) %>% summarise(interval_average = round(mean(steps, na.rm=TRUE)))
for(row in 1:nrow(replace_data)){
  if(is.na(replace_data$steps[row])){
    replace_data$steps[row] <- interval_averages[interval_averages$interval==replace_data$interval[row],]$interval_average
  }
}

replace_summary <- replace_data %>% select(steps, date) %>% group_by(date) %>% summarise(total_steps = sum(steps, na.rm = TRUE))
hist(replace_summary$total_steps, main = "Histogram of Total Steps/day without Missing Data", xlab = "Total daily steps")

replace_mean <- mean(replace_summary$total_steps)
replace_median <- median(replace_summary$total_steps)

```
There are `r format(total_missing, scientific = FALSE)` missing values in the data set. The missing data was relaced with the average for the interval with the missing data. The new avarage is `r format(replace_mean, scientific = FALSE)`, this is a change of `r round(replace_mean - daily_mean, digits = 2) ` steps. The new median is `r format(replace_median, scientific = FALSE, digits = 2)`, this is a change of `r replace_median - daily_median` steps.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday}
weekday_labels <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activity_data$weekday <- factor(weekdays(as.Date(as.character(activity_data$date))) %in% weekday_labels, levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))

weekday <- activity_data[activity_data$weekday == "weekday",]
weekend <- activity_data[activity_data$weekday == "weekend",]

weekday_summary <- weekday %>% select(steps, date, interval) %>% group_by(interval) %>% summarise(average_steps = mean(steps, na.rm=TRUE))

weekend_summary <- weekend %>% select(steps, date, interval) %>% group_by(interval) %>% summarise(average_steps = mean(steps, na.rm=TRUE))

par( mfrow = c(2,1))
par(mar = c(0,0,0,0), oma = c(4,4,.5,.5))
plot(weekday_summary$interval, weekday_summary$average_steps, type = 'l', xaxt = 'n',ann = FALSE, ylim = c(0,250))
mtext("Weekday", side = 3, adj = 0.05, line = -1.3)
plot(weekend_summary$interval, weekend_summary$average_steps, type = 'l', ylim = c(0,250), xlab = "Interval")
mtext("Weekend", side = 3, adj = 0.05, line = -1.3)
mtext("Steps", 2, 3, outer=TRUE, las=0)






